---
title: "clean_data"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

# 0) NOTES

Purpose of document is to clean dataset of California ballot measures, and create basic visualization and statistics.


**********************************************************

# 1) SET UP 

## 1.1. General 

```{r}
# remove all objects
rm(list=ls())

# Unload all packages 
library(pacman)
pacman::p_unload(all)

# Add packages 
pacman::p_load(
  tidyverse, #dplyr, readr, etc.
  data.table, #fread() 
  foreign, #load data types including stata .dta files 
  magrittr, #%<>% operator
  skimr, #for summerising
  lubridate, #dates
  stargazer, #pretty tables 
  xtable #pretty tables
)
```

## 1.2. Load Data 

```{r}
    # Set the working directory
setwd("/home/dviryo/CompClassFall2019/assignments/prop/analsys")
getwd()
    # load data
load("/home/dviryo/CompClassFall2019/assignments/prop/data/cal_propall_data.RData")

```

**********************************************************

# 2) CLEAN DATA

## 2.1. Check Data 

```{r}
cal_prop_all %>% skimr::skim()
```

```{r}
# Trim whitespace from ends of all character variables 
cal_prop_all %<>%
  mutate_if(is.character, list(str_trim))

# Check for empty cells 
cal_prop_all %>%
  # Keep only character variables 
  select_if(is.character) %>%
  # Recode character variables as 1 if cell is empty 
  mutate_all(list(~ifelse(.=="",1,0))) %>%
  # Add up empty cells for each character variable 
  summarise_all(sum, na.rm=T) %>%
  # Transpose data for visibility 
  t() 

# Remove duplicate rows based on all columns
cal_prop_all %<>%
  distinct()

```
 

## 2.2. Gen vars with diff types 
 
```{r}
cal_prop_all %<>%
  # Create binary variable for status 
  mutate(status = ifelse(grepl("Pass", status), T, F)) %>%
  # Convert percent to numeric type (removing unofficial results)
  mutate(percent = str_remove(percent, "%")) %>%
  mutate(percent = as.numeric(percent)) %>%
  # assign year as numeric and all others as char
  mutate(year = as.numeric(year)) %>%
  mutate(topics = as.character(topics)) %>%
  mutate(title = as.character(title)) %>%
  mutate(election = as.character(election)) %>%
  mutate(type = as.character(type))
``` 
```{r}
cal_prop_all %>% skimr::skim()
```

## 2.3 Seperate measures by topics and add variables that measure overwhelming success/failure

```{r}
  # for every measure with more than 1 topic - duplicate for each topic
for (i in 1:nrow(cal_prop_all)) {
  for (j in 8:12) {
    if (!is.na(cal_prop_all[i,j])){
    cal_prop_all %<>%
      add_row(title = as.character(cal_prop_all[i,1]), election = as.character(cal_prop_all[i,2]), year = as.numeric(cal_prop_all[i,3]), type = as.character(cal_prop_all[i,4]), status = as.logical(cal_prop_all[i,5]), percent = as.numeric(cal_prop_all[i,6]), topics = as.character(cal_prop_all[i,j]))
    }
  }
}

# Add "polar" variable to measure level of concesus
cal_prop_all %<>%
  mutate(polar = NA) %>%
  mutate(polar = ifelse(percent > 30 & percent < 70, "0", polar)) %>%
  mutate(polar = ifelse(percent <= 30, "1", polar)) %>%
  mutate(polar = ifelse(percent >= 70, "2", polar)) %>%
  mutate(polar = as.numeric(polar))

# Add "veryS" variable to indicate an overwhelmingly successful measure
cal_prop_all %<>%
  mutate(veryS = NA) %>%
  mutate(veryS = ifelse(percent >= 70, 1, veryS))

#Add "veryUnS" variable to indicate an overwhelmingly un-successful measure
cal_prop_all %<>%
  mutate(veryUnS = NA) %>%
  mutate(veryUnS = ifelse(percent <= 30, 1, veryUnS))
```



# 3) SUMMARISE DATA

```{r}
cal_prop_all %>% glimpse()

# Symmary Stats
summary_stats <- cal_prop_all %>% 
  group_by(as.character(topics)) %>%
  summarise(avgPercentYesVote=mean(percent, na.rm=T),
            sdPercentYes=sd(percent, na.rm=T),
            medPercentYes=median(percent, na.rm=T),
            observations=n(),
            PercentVeryS=sum(veryS, na.rm=T) / observations,
            PercentVeryUnS=sum(veryUnS, na.rm=T) / observations,
            PercentPass=sum(as.logical(status)) / observations)
            
#avgPercentYesVote - higher means more chances for future success.
#sdPercentYes - higher means less consistent public opnion.
#medPercentYes - higher means more chances for future success.
#observations
#PercentVeryS - percent of total that passed with overwhelmingly success (higher than 70% yes vote)
#PercentVeryUnyS - percent of total that overwhelmingly failed (less than 30% yes votes)
#percentpass - percent of total that passed
summary_stats
cal_prop_all %>% skimr::skim()
```


**********************************************************

# 4) SAVE DATA

```{r}
save(cal_prop_all, file="../data/cal_prop_all_clean.RData")
```

# 5) visualization
```{r}
 # load data
load("/home/dviryo/CompClassFall2019/assignments/prop/data/cal_prop_all_clean.RData")
```

## 5.1 ditch all the measures with less than 60 observations
```{r}
  #keep only topics with more than 60 observations
high_observations <- c()
for (k in 1:nrow(summary_stats)) {
  if (summary_stats[k,5] > 60) {
    high_observations <- c(high_observations,as.character(summary_stats[k,1]))
    k = k + 1
  }
}
# create a dataframe with only high observation topics
cal_prop_greater60 <- filter(cal_prop_all, topics %in% high_observations)
skim(cal_prop_greater60)
# create summary stats for the new dataframe, with an additional grouping by type - iniative/referendum
summary_stats_greater60 <- cal_prop_greater60 %>% 
  group_by(topics, type) %>%
  summarise(avgPercentYesVote=mean(percent, na.rm=T),
            sdPercentYes=sd(percent, na.rm=T),
            medPercentYes=median(percent, na.rm=T),
            observations=n(),
            PercentVeryS=sum(veryS, na.rm=T) / observations,
            PercentVeryUnS=sum(veryUnS, na.rm=T) / observations,
            PercentPass=sum(as.logical(status)) / observations)

summary_stats_greater60
```

## 5.2 compare the overall pass/fail
```{r} 
cal_prop_greater60 %>% arrange(status) %>%
ggplot(aes(x = reorder(topics, -status), y= year, fill = as.factor(status))) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous("Proportion") +
  scale_x_discrete("", expand=c(0, 0)) +
  #scale_fill_identity()
  coord_flip()+
  scale_fill_discrete(name = "", labels = c("Fail", "Pass"))

ggsave("pass_vs_fail.png")
```

## 5.3 how much did the pass rate changed over time? what is the future prediction?
```{r}
#change in PASS rate over time
ggplot(data = cal_prop_greater60, aes(x = year, y = as.numeric(status), color=topics)) +
  geom_point(size=0.5)+
  stat_smooth(method = "glm", 
                method.args = list(family = "binomial"), formula = y ~ splines::ns(x, 1)) +
  xlab("year")+
  ylab("pass/fail")+
  theme_bw()+
  facet_wrap( ~ topics)

ggsave("change_over_time.png")
```

## 5.4 are there differences in the pass rate between iniative/legislative referendum/popular referendum?
```{r}
#ggplot bar per type
ggplot(data=cal_prop_greater60, aes(x = type, y = year, fill = as.factor(status))) +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous("Proportion") +
  scale_x_discrete("", expand=c(0, 0)) +
  facet_wrap( ~ topics) +
  scale_fill_discrete(name = "", labels = c("Fail", "Pass"))

ggsave("iniative_vs_referendum.png")
```
```{r}
#ggplot line per type
ggplot(data=summary_stats_greater60, aes(x = reorder(topics, -PercentPass), y = PercentPass, colour = topics == "Criminal Justice")) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
  geom_point() +
  facet_wrap( ~ type) +
  #scale_fill_discrete(name = "", labels = c("Fail", "Pass"))

ggsave("iniative_vs_referendum2.png")
```

## 5.5 NOT RUNNIG cool/harmless point visualisation - difference in overwhelming success/fail
```{r}
#drop observations with NA in the percent column
cal_prop_greater60 %<>% 
  arrange(desc(percent)) %>%
  drop_na(percent)
```

## 5.6 compare the overwhelming success/fail
```{r} 
ggplot(cal_prop_greater60, aes(x = reorder(topics, -polar), y= percent, fill = as.factor(polar))) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous("Proportion") +
  scale_x_discrete("", expand=c(0, 0)) +
  #scale_fill_identity()
  coord_flip() +
  scale_fill_discrete(name = "Percent Yes", labels = c("30%-60%", "Less than 30%", "More than 70%"))
  
ggsave("overwhelming_compare.png")
```


